Forecasting Gamma-ray Bursts using Gravitational Waves

### 2018-04-02 Paper/Manuscript.tex
Updates

A necessary, but insufficient, condition for tidal disruption (TD) is that
\be
\f{2 G M_\text{BH} (c_R R_\text{NS})}{r^3} \gtrsim \f{G M_\text{NS}}{(c_R R_\text{NS})^2} \label{eq:TD_cond1},
\ee
where $r$ is the orbital radius, $R_\text{NS}$ is the radius of the NS and $M$ denotes masses. 
$c_R R_\text{NS}$ is the semi-major axis of the elongated oblate spheroid that represents the tidally distorted shaped of the NS to leading order with
$c_R >1 $. %Setting $m_2= M_\text{BH}, m_1 = M_\text{NS}$, and $r_1=r_\text{NS}$, 
The above conditions roughly implies
%
\be
\f{M_\text{BH}}{M_\text{NS}} \gtrsim \left(\f{r}{R_\text{NS}}\right)^3 \label{eq:TD_cond2}.
\ee
%
We wish to maximize the chances of observing a TD event, therefore we ingrain the BH-NS binary with certain desirable features
some of which can be deduced from Eqs.~(\ref{eq:TD_cond1}, \ref{eq:TD_cond2}). These are
%
%
\begin{enumerate}[(i)]
 \item Small separation. As the Newtonian tidal force due to the black hole is proportional to $M_\text{BH}/r^3$, $r$ has to be
 minimized to provide maximum tidal force. 
 However, for $r< r_\text{ISCO}$ the NS simply plunges into the BH; therefore the minium separation is given by the ISCO radius.
 \item Low black hole mass. In the geodesic case, $r_\text{ISCO} \propto M_\text{BH}$, so the Newtonian tidal force roughly scales as $M_\text{BH}/r_\text{ISCO}^3\propto M_\text{BH}^{-2}$. 
 Hence, $M_\text{BH}$ must also be minimized. 
 As was the case with $r$, there is also a lower bound to $M_\text{BH}$ set from astrophysical observations and stellar evolution models, which currently yield $M_\text{BH} \gtrsim 5 M_\odot$ \cite{Farr:2010tu, Raithel:2017nlc, Ozel:2010su, Wiktorowicz:2013dua, Christian:2018mjv}.
 \item High black hole spins. $r_\text{ISCO}$ is much smaller in the case of a prograde NS orbit around a spinning Kerr black hole.
 For instance, for the maximum dimensionless BH spin value of $a\equiv c J_\text{BH}/(G M^2_\text{BH})=1$, $r_\text{ISCO}$ for prograde orbits is 1/6th of the Schwarzschild value ($6GM_\text{BH}/c^2$) whereas for $a=1$ retrograde orbits, $r_\text{ISCO}$ is 3/2 of the Schwarzschild value.
 With smaller ISCO separation, there is much more tidal force exerted on the NS before it plunges in.
 \item Less compact neutron stars. Eq.~(\ref{eq:TD_cond1}) can be rearranged to isolate $c_\text{NS}$
 on the right hand side. Smaller values for $c_\text{NS}$ mean that TD is more likely to occur for $r>r_\text{ISCO}$.
\end{enumerate}
%
We are thus led to set $M_\text{BH} =5 M_\odot$. There is additional motivation to choose low BH mass as 
 $\tau_\text{insp} \propto M_c^{-5/3} = (1+Q)^{1/3}/Q\to Q^{-2/3}$ as $Q\to\infty$ for $Q\equiv m_2/m_1 > 1$, i.e., heavier binaries inspiral faster.
 We once again set $M_\text{NS}=1.4 M_\odot$ thus $Q=5/1.4  \simeq 3.571$. $c_\text{NS}$ is determined by the radius of the NS constrained in the $9.9 - 11.2\,$km range \cite{Ozel:2016oaf}
 which results in $0.13\lesssim c_\text{NS} \lesssim 0.15$ 
 for our chosen NS mass.
%As we wish to avoid discussing the details of the strong-field evolution of the binary, we borrow certain results from the literature to
%justify our choice for the chosen parameters above. 
If we pick the more pessimistic $c_\text{NS} = 0.15$ bound then we must have $Q\lesssim 5$ and $a\gtrsim 0.5$ (prograde) for TD to occur \cite{Ferrari:2008nr}
and $Q< 3$ in the case of $a=0$ \cite{Kyutoku:2011vz}.

Although the results from each numerical study may somewhat vary what is clear is that we must have high spin and low black hole mass.
This raises an interesting astrophysical question as to whether it is possible to have such light-weight high-spin black holes assuming that
they are spun up by accretion from their secondary (that will later on become a NS) in binary systems, thus they will have to gain mass.
It is possible that the mass gain pushes $Q$ too high to either cause a direct plunge of the NS before TD or decrease $\tau_\text{insp}$ considerably
to eliminate any chances of getting an advance warning.
However, if the spin of the BH comes from the supernova explosion mechanism itself as argued for G1915 in Ref.~\cite{McClintock:2006xd} then it is possible to have a 
low-mass BH in a binary system.

Therefore, we simply assume that such lightweight, high-spin BHs exist in BH-NS binaries and compute strains and SNRs accordingly.
A quick calculation shows that these systems yield $T_\text{AW} \lesssim 30\,$seconds in the ALV network of 2020s. %in Table \ref{table:LIGO2020}is now less than $ 40\,$ seconds. 
Therefore, we simply focus on the capabilities of ET which we summarize in Table~\ref{table:ET_BH_NS} below.
%
%
%
\begin{table}[h]
\centering
\begin{tabular}{l|ccccc}
%toprule
\hline
$D\,$(Mpc) & \multicolumn{2}{c}{ET-B} &  & \multicolumn{2}{c}{ET-C}\T\B\\
\hline
{}&  \ \hspace{6mm}$T_\text{AW}$ \hspace{8mm} & $\bar{\rho}_{F}$ &{}  & $\hspace{6mm}T_\text{AW} \hspace{10mm}$ & $\bar{\rho}_{F}$\T\B\\
%midrule

>>>>

A necessary, but insufficient, condition for tidal disruption (TD) is that
\be
\f{2 G M_\text{BH} (c_R R_\text{NS})}{r^3} \gtrsim \f{G M_\text{NS}}{(c_R R_\text{NS})^2} \label{eq:TD_cond1},
\ee
where $r$ is the orbital radius, $R_\text{NS}$ is the radius of the NS and $M$ denotes masses. 
$c_R R_\text{NS}$ is the semi-major axis of the elongated oblate spheroid that represents the tidally distorted shaped of the NS to leading order with
$c_R >1 $. %Setting $m_2= M_\text{BH}, m_1 = M_\text{NS}$, and $r_1=r_\text{NS}$, 
The above conditions roughly implies
%
\be
\f{M_\text{BH}}{M_\text{NS}} \gtrsim \left(\f{r}{R_\text{NS}}\right)^3 \label{eq:TD_cond2}.
\ee
%
We wish to maximize the chances of observing a TD event, therefore we ingrain the BH-NS binary with certain desirable features
some of which can be deduced from Eqs.~(\ref{eq:TD_cond1}, \ref{eq:TD_cond2}). These are
%
%
\begin{enumerate}[(i)]
 \item Small separation. As the Newtonian tidal force due to the black hole is proportional to $M_\text{BH}/r^3$, $r$ has to be
 minimized to provide maximum tidal force. 
 However, for $r< r_\text{ISCO}$ the NS simply plunges into the BH; therefore the minium separation is given by the ISCO radius.
 \item Low black hole mass. In the geodesic case, $r_\text{ISCO} \propto M_\text{BH}$, so the Newtonian tidal force roughly scales as $M_\text{BH}/r_\text{ISCO}^3\propto M_\text{BH}^{-2}$. 
 Hence, $M_\text{BH}$ must also be minimized. 
 As was the case with $r$, there is also a lower bound to $M_\text{BH}$ set from astrophysical observations and stellar evolution models, which currently yield $M_\text{BH} \gtrsim 5 M_\odot$ \cite{Farr:2010tu, Raithel:2017nlc, Ozel:2010su, Wiktorowicz:2013dua, Christian:2018mjv}.
 \item High black hole spins. $r_\text{ISCO}$ is much smaller in the case of a prograde NS orbit around a spinning Kerr black hole.
 For instance, for the maximum dimensionless BH spin value of $a\equiv c J_\text{BH}/(G M^2_\text{BH})=1$, $r_\text{ISCO}$ for prograde orbits is 1/6th of the Schwarzschild value ($6GM_\text{BH}/c^2$) whereas for $a=1$ retrograde orbits, $r_\text{ISCO}$ is 3/2 of the Schwarzschild value.
 With smaller ISCO separation, there is much more tidal force exerted on the NS before it plunges in.
 \item Less compact neutron stars. Eq.~(\ref{eq:TD_cond1}) can be rearranged to isolate $c_\text{NS}$
 on the right hand side. Smaller values for $c_\text{NS}$ mean that TD is more likely to occur for $r>r_\text{ISCO}$.
\end{enumerate}
%

Although the results from each numerical study may somewhat vary what is clear is that we must have high spin and low black hole mass. We are thus led to set $M_\text{BH} =5 M_\odot$. 
This raises an interesting astrophysical question as to whether it is possible to have such light-weight high-spin black holes assuming that
they are spun up by accretion from their secondary (that will later on become a NS) in binary systems, thus they will have to gain mass.
It is possible that the mass gain pushes $M_\text{BH}$ too high to either cause a direct plunge of the NS before TD or decrease $\tau_\text{insp}$ considerably
to eliminate any chances of getting an advance warning.
However, if the spin of the BH comes from the supernova explosion mechanism itself as argued for G1915 in Ref.~\cite{McClintock:2006xd} then it is possible to have a 
low-mass BH in a binary system\footnote{We thank Dr. Morgan Fraser
for pointing this out.}.

There is additional motivation to choose low BH mass as 
 $\tau_\text{insp} \propto M_c^{-5/3} = (1+Q)^{1/3}/Q$ for $Q\equiv M_\text{BH}/M_text{NS} > 1$, which scales as $Q^{-2/3}$ as $Q\to\infty$ , i.e., heavier binaries inspiral faster.
 We once again set $M_\text{NS}=1.4 M_\odot$ giving $Q=5/1.4  \simeq 3.571$. $c_\text{NS}$ is determined by the radius of the NS constrained in the $9.9 - 11.2\,$km range \cite{Ozel:2016oaf}
 which results in $0.13\lesssim c_\text{NS} \lesssim 0.15$ 
 for our chosen NS mass.
%As we wish to avoid discussing the details of the strong-field evolution of the binary, we borrow certain results from the literature to
%justify our choice for the chosen parameters above. 
If we pick the more pessimistic $c_\text{NS} = 0.15$ bound then we must have $Q\lesssim 5$ and $a\gtrsim 0.5$ (prograde) for TD to occur \cite{Ferrari:2008nr}
and $Q< 3$ in the case of $a=0$ \cite{Kyutoku:2011vz}.

So, we simply assume that such lightweight, high-spin BHs exist in BH-NS binaries and compute strains and SNRs accordingly.
A quick calculation shows that these systems yield $T_\text{AW} \lesssim 30\,$seconds in the ALV network of 2020s. %in Table \ref{table:LIGO2020}is now less than $ 40\,$ seconds. 
Therefore, we simply focus on the capabilities of ET which we summarize in Table~\ref{table:ET_BH_NS} below.
%
%
%
\begin{table}[h]
\centering
\begin{tabular}{l|ccccc}
%toprule
\hline
$D\,$(Mpc) & \multicolumn{2}{c}{ET-B} &  & \multicolumn{2}{c}{ET-C}\T\B\\
\hline
{}&  \ \hspace{6mm}$T_\text{AW}$ \hspace{8mm} & $\bar{\rho}_{F}$ &{}  & $\hspace{6mm}T_\text{AW} \hspace{10mm}$ & $\bar{\rho}_{F}$\T\B\\
%midrule
